A Comparison of R Tools for Nonlinear Least Squares Modeling

### 2024-02-08 copyedit240208/20230-15-copyeditresponse.Rmd
copyedit temp

---
title: "A Comparison of R Tools for Nonlinear Least Squares Modeling"
date:  \`r format(Sys.Date(), "%B %d, %Y")`\
abstract: >
  Our Google Summer of Code project "Improvements to `nls()`" considered the 
  features and limitations of the R function `nls()` with the aim 
  of improving and rationalizing R tools for nonlinear regression.
  The rich features of `nls()` are weakened by several deficiencies and
  inconsistencies such as a lack of stabilization of the Gauss-Newton solver.
  Further considerations are the usability and  maintainability of
  the code base that provides the functionality `nls()` claims to offer.
  Various packages, including our `nlsr`, provide alternative capabilities.
  We consider the differences in goals, approaches, and features of
  different tools for nonlinear least squares modeling in R. Discussion of 
  these matters is relevant to improving R generally as well as its 
  nonlinear estimation tools.

@@

`nls()` is the tool for estimating nonlinear statistical models in base R, the 
primary software distribution
from the Comprehensive R Archive Network (*https://cran.r-project.org*).
The function dates to the 1980s and the work related to @bateswatts in S (see 
*https://en.wikipedia.org/wiki/S_%28programming_language%29*).

The `nls()` function has a remarkable and comprehensive set of capabilities for 
estimating nonlinear models that are expressed as **formulas**. In particular, we
note that it

- handles formulas that include R functions, even ones that call calculations in
  other programming languages
- allows data to be weighted or subset
- can estimate bound-constrained parameters
- provides a mechanism for handling partially linear models
- permits parameters to be indexed over a set of related data
- produces measures of variability (i.e., standard error estimates) for the 
estimated parameters
- has related profiling capabilities for exploring the likelihood surface as 
parameters are changed
- links to many pre-coded (`selfStart`) models that do not require
  initial parameter values.

With such a range of features and long history, the code has become untidy 
and overly patched, difficult to maintain, and its underlying methods could
be improved. Various workers have developed packages to overcome these 
concerns, and we will address some of these here.

@@

The tools we will consider are the base-R `nls()` function 
and some packages that are available on the CRAN repository.
Of these, we will pay particular attention to **nlsr** (@nlsr2023manual),
**minpack.lm** (@minpacklm12), and **gslnls** (@gslnls23) which are
general nonlinear least-squares solvers.

While we will provide capsule comments for some other CRAN packages, 
we will not pursue those in the Bioconductor (@Bioconductor) 
collection, nor those on repositories such as 
GitHub (https://github.com) and Gitlab (https://about.gitlab.com).

@@

Our work was aimed at unifying nonlinear modeling functionality in R,
ideally in a refactored `nls()` function.

For R **users**, we would advise adapting working scripts, preferably those
with documentation and using recent tools. If there is a suspicion that there
may be ill-conditioning, package `nlsr` or the example we give below of how to
find singular values of the Jacobian allow these diagnostics to be calculated.

For R **developers**, we invite and encourage discussion of the design choices, since these
have downstream implications for ease of use, adaptation to new features, and 
efficiency of ongoing maintenance.

>>>>

---
title: "A Comparison of R Tools for Nonlinear Least Squares Modeling"
date:  \`r format(Sys.Date(), "%B %d, %Y")`\
abstract: >
  Our Google Summer of Code project "Improvements to `nls()`" investigated
  rationalizing R tools for nonlinear regression and nonlinear estimation tools by
  considering usability, maintainability and functionality, especially for a
  Gauss-Newton solver.
  <!-- JN: copyedit essentially accepted -->
  The rich features of `nls()` are weakened by several deficiencies and
  inconsistencies such as a lack of stabilization of the Gauss-Newton solver.
  Further considerations are the usability and  maintainability of
  the code base that provides the functionality `nls()` claims to offer.
  Various packages, including our `nlsr`, provide alternative capabilities.
  We consider the differences in goals, approaches, and features of
  different tools for nonlinear least squares modeling in R. Discussion of 
  these matters is relevant to improving R generally as well as its 
  nonlinear estimation tools.

@@

`nls()` is a Comprehensive R Archive Network (CRAN: *https://cran.r-project.org*) 
tool that has remarkable and wide-ranging features for estimating nonlinear
statistical models that are expressed as **formulas**. 
<!-- JN: copyedit comment integrated -->
The function dates to the 1980s and the work related to @bateswatts in S (see 
*https://en.wikipedia.org/wiki/S_%28programming_language%29*).

In particular, we
note that it

- can call calculations in other programming languages
<!-- JN: copyedit comment integrated -->
- allows weighted or subset data
<!-- JN: copyedit comment integrated -->
- can estimate bound-constrained parameters
- provides partially linear model handling mechanism
<!-- JN: copyedit comment integrated -->
- permits parameters to be indexed over a set of related data
- produces measures of variability (i.e., standard error estimates) for the 
estimated parameters
- has related profiling capabilities for exploring the likelihood surface as 
parameters are changed
- links to many pre-coded (`selfStart`) models that do not require
  initial parameter values.

Due to its extensive range of features and prolonged history, the code has become untidy and overly patched, making it challenging to maintain and ripe for improvement in its underlying methods. 
<!-- JN: copyedit comment integrated -->

@@

Besides the base-R `nls()` function, we will pay particular attention 
to **nlsr** (@nlsr2023manual),
**minpack.lm** (@minpacklm12), and **gslnls** (@gslnls23) which are
general nonlinear least-squares solvers in the CRAN repository.
While we will provide capsule comments for some other CRAN packages, 
those in the Bioconductor (@Bioconductor) collection are more
specialized, and those on repositories such as 
GitHub (https://github.com) and Gitlab (https://about.gitlab.com),
while interesting, do not have the checking applied to CRAN packages.
<!-- JN: revised text combining 2 copyedit comments -->

@@

Our work aimed at unifying nonlinear modeling functionality in R,
<!-- JN: removed "was" as suggesgted -->
ideally in a refactored `nls()` function.

For R **users**, we would advise adapting working scripts, preferably those
with documentation and using recent tools. If there is a suspicion that there
may be ill-conditioning, package `nlsr` or the example we give in the 
section "Comparison notes for formula-setup solutions" below of how to
find singular values of the Jacobian allow these diagnostics to be calculated.
<!-- JN: reference to section added ??have I done this correctly? -->

For R **developers**, we invite and encourage discussion of the design choices, 
since these
have downstream implications for ease of use, adaptation to new features, and 
efficiency of ongoing maintenance.
<!-- JN: Could not understand how copyedit "such as" would be applied -->

### 2024-02-08 copyedit240208/20230-15-copyeditresponse.tex
copyedit temp

\abstract{%
Our Google Summer of Code project ``Improvements to \texttt{nls()}'' considered the features and limitations of the R function \texttt{nls()} with the aim of improving and rationalizing R tools for nonlinear regression. The rich features of \texttt{nls()} are weakened by several deficiencies and inconsistencies such as a lack of stabilization of the Gauss-Newton solver. Further considerations are the usability and maintainability of the code base that provides the functionality \texttt{nls()} claims to offer. Various packages, including our \texttt{nlsr}, provide alternative capabilities. We consider the differences in goals, approaches, and features of different tools for nonlinear least squares modeling in R. Discussion of these matters is relevant to improving R generally as well as its nonlinear estimation tools.
}

@@

\texttt{nls()} is the tool for estimating nonlinear statistical models in base R, the
primary software distribution
from the Comprehensive R Archive Network (\emph{\url{https://cran.r-project.org}}).
The function dates to the 1980s and the work related to D. M. Bates and Watts (1988) in S (see
\emph{\url{https://en.wikipedia.org/wiki/S_\%28programming_language\%29}}).

The \texttt{nls()} function has a remarkable and comprehensive set of capabilities for
estimating nonlinear models that are expressed as \textbf{formulas}. In particular, we
note that it

\begin{itemize}
\tightlist
\item
  handles formulas that include R functions, even ones that call calculations in
  other programming languages
\item
  allows data to be weighted or subset
\item
  can estimate bound-constrained parameters
\item
  provides a mechanism for handling partially linear models
\item
  permits parameters to be indexed over a set of related data
\item
  produces measures of variability (i.e., standard error estimates) for the
  estimated parameters
\item
  has related profiling capabilities for exploring the likelihood surface as
  parameters are changed
\item
  links to many pre-coded (\texttt{selfStart}) models that do not require
  initial parameter values.
\end{itemize}

With such a range of features and long history, the code has become untidy
and overly patched, difficult to maintain, and its underlying methods could
be improved. Various workers have developed packages to overcome these
concerns, and we will address some of these here.

@@

The tools we will consider are the base-R \texttt{nls()} function
and some packages that are available on the CRAN repository.
Of these, we will pay particular attention to \textbf{nlsr} (John C. Nash and Murdoch (2023)),
\textbf{minpack.lm} (Elzhov et al. (2012)), and \textbf{gslnls} (Chau (2023)) which are
general nonlinear least-squares solvers.

While we will provide capsule comments for some other CRAN packages,
we will not pursue those in the Bioconductor (Gentleman et al. (2004))
collection, nor those on repositories such as
GitHub (\url{https://github.com}) and Gitlab (\url{https://about.gitlab.com}).

@@

Our work was aimed at unifying nonlinear modeling functionality in R,
ideally in a refactored \texttt{nls()} function.

For R \textbf{users}, we would advise adapting working scripts, preferably those
with documentation and using recent tools. If there is a suspicion that there
may be ill-conditioning, package \texttt{nlsr} or the example we give below of how to
find singular values of the Jacobian allow these diagnostics to be calculated.

@@

\includegraphics{ComparingNLStools_files/figure-latex/ex01-1.pdf}

@@

\begin{verbatim}
#> tgslnls1  -- ss= 9205.4 :  Asym = 35.532  xmid = 20846  scal = -1741.1; 47  itns
\end{verbatim}

@@

\begin{verbatim}
#> [1] 3.4641e+00 9.4495e-09 4.1324e-11
\end{verbatim}

>>>>

\abstract{%
Our Google Summer of Code project ``Improvements to \texttt{nls()}'' investigated rationalizing R tools for nonlinear regression and nonlinear estimation tools by considering usability, maintainability and functionality, especially for a Gauss-Newton solver. The rich features of \texttt{nls()} are weakened by several deficiencies and inconsistencies such as a lack of stabilization of the Gauss-Newton solver. Further considerations are the usability and maintainability of the code base that provides the functionality \texttt{nls()} claims to offer. Various packages, including our \texttt{nlsr}, provide alternative capabilities. We consider the differences in goals, approaches, and features of different tools for nonlinear least squares modeling in R. Discussion of these matters is relevant to improving R generally as well as its nonlinear estimation tools.
}

@@

\texttt{nls()} is a Comprehensive R Archive Network (CRAN: \emph{\url{https://cran.r-project.org}})
tool that has remarkable and wide-ranging features for estimating nonlinear
statistical models that are expressed as \textbf{formulas}.
The function dates to the 1980s and the work related to D. M. Bates and Watts (1988) in S (see
\emph{\url{https://en.wikipedia.org/wiki/S_\%28programming_language\%29}}).

In particular, we
note that it

\begin{itemize}
\tightlist
\item
  can call calculations in other programming languages
\item
  allows weighted or subset data
\item
  can estimate bound-constrained parameters
\item
  provides partially linear model handling mechanism
\item
  permits parameters to be indexed over a set of related data
\item
  produces measures of variability (i.e., standard error estimates) for the
  estimated parameters
\item
  has related profiling capabilities for exploring the likelihood surface as
  parameters are changed
\item
  links to many pre-coded (\texttt{selfStart}) models that do not require
  initial parameter values.
\end{itemize}

Due to its extensive range of features and prolonged history, the code has become untidy and overly patched, making it challenging to maintain and ripe for improvement in its underlying methods.

@@

Besides the base-R \texttt{nls()} function, we will pay particular attention
to \textbf{nlsr} (John C. Nash and Murdoch (2023)),
\textbf{minpack.lm} (Elzhov et al. (2012)), and \textbf{gslnls} (Chau (2023)) which are
general nonlinear least-squares solvers in the CRAN repository.
While we will provide capsule comments for some other CRAN packages,
those in the Bioconductor (Gentleman et al. (2004)) collection are more
specialized, and those on repositories such as
GitHub (\url{https://github.com}) and Gitlab (\url{https://about.gitlab.com}),
while interesting, do not have the checking applied to CRAN packages.

@@

Our work aimed at unifying nonlinear modeling functionality in R,
ideally in a refactored \texttt{nls()} function.

For R \textbf{users}, we would advise adapting working scripts, preferably those
with documentation and using recent tools. If there is a suspicion that there
may be ill-conditioning, package \texttt{nlsr} or the example we give in the
section ``Comparison notes for formula-setup solutions'' below of how to
find singular values of the Jacobian allow these diagnostics to be calculated.

@@

\includegraphics{20230-15-copyeditresponse_files/figure-latex/ex01-1.pdf}

@@

\begin{verbatim}
#> tgslnls1  -- ss= 9205.4 :  Asym = 35.532  xmid = 20846  scal = -1740.4; 47  itns
\end{verbatim}

@@

\begin{verbatim}
#> [1] 3.4641e+00 1.0020e-08 2.9378e-11
\end{verbatim}
